### 2019-11-27 paper/paper.tex
Add pmerge and other updates from Overleaf

In $N (k \mathbin{:} 1) \mathbin{:} M$ scheduling, we have $N$ units of work mapped onto $M$ CPU threads. Additionally, each of those units of work may be composed of $k$ cooperatively scheduled tasks. This is achieved by pinning the $k$ tasks in a group to one CPU thread, while load balancing units of work across the available cores using the \verb|partr| scheduler, a novel implementation of a parallel depth-first scheduling algorithm.

@@

\begin{lstlisting}[language = Julia]
# perform a merge sort on `v` using parallel threads
function psort(v::AbstractVector)
    hi = length(v)
    if hi < 100_000 # below some cutoff, run in serial
        return sort(view(v, 1:hi), alg = MergeSort)
    end
# split the range and sort the halves in parallel recursively
    mid = (1 + hi) >>> 1
    half = Threads.@spawn psort(view(v, 1:mid))
    right = psort(view(v, (mid + 1):hi))
    left = fetch(half)::typeof(right)
# perform the merge on the result
    out = similar(v)
    i, il, ir = 1, 1, 1
    @inbounds while il <= mid && ir <= (hi - mid)
        l, r = left[il], right[ir]
        if r < l
            out[i] = r
            ir += 1
        else
            out[i] = l
            il += 1
        end
        i += 1
    end
    @inbounds while il <= length(left)
        out[i] = left[il]
        il += 1
        i += 1
    end
    @inbounds while ir <= length(right)
        out[i] = right[ir]
        ir += 1
        i += 1
    end
    return out
end
\end{lstlisting}

@@

@@

\begin{lstlisting}[language = Julia, inputencoding=utf8, literate = {⊕}{{$\oplus$}}1]
using .Threads: @threads
function prefix_threads!(⊕, y::AbstractVector)
    l = length(y)
    k = ceil(Int, log2(l))
    # do reduce phase
    for j = 1:k
        @threads for i = 2^j:2^j:min(l, 2^k)
            @inbounds y[i] = y[i - 2^(j - 1)] ⊕ y[i]
        end
    end
    # do expand phase
    for j = (k - 1):-1:1
        @threads for i = 3*2^(j - 1):2^j:min(l, 2^k)
            @inbounds y[i] = y[i - 2^(j - 1)] ⊕ y[i]
        end
    end
    return y
end

@@

Changing the I/O code (files, streams, folders, and other platform code) to work on from any thread was another big project. The existing design requires an underlying asynchronous library, with a design similar to Windows IOCP, to efficiently manage large numbers of open event sources and provide the simplicity and concision of the logic shown in \ref{subsub:background} on all platforms. For this, we have been using the libuv library. This lets us have most platform-specific code isolated in a separately tested library and provide more commonality in our runtime library. As an initial implementation to make this library safe to use from threads, we've used one big IO lock around all calls to it. However, this library also has callbacks and will block to wait for external incoming events, so we also needed to integrate it fully with the task scheduler to get it to cooperatively release the lock on demand. We were able to do so by adding an asynchronous channel (\verb|uv_async_t|) to wake the one thread running the event loop while all other threads sleep on a system mutex (\verb|uv_cond_t|) when there is no work for them to perform. When try entering the event loop, we do so only if the count of currently waiting tasks is zero. In the future, this work may allow us to move the event loop entirely to a separate thread (and/or multiple threads). It seems that this design change may thus be making threading support a mandatory requirement for the underlying VM—with the advantage we that we can get more throughput on the large-core systems that are only becoming more common.

@@

A prototype implementation of the \verb|partr| scheduler was first written for us in C by Kiran Pamnany of Intel back in late 2016, following research done on scheduling threads for beneficial cache sharing for best throughput\cite{Chen:2007:STC:1248377.1248396}. The goal of this work was effortless composition of threading-capable libraries with a globally depth-first work ordering (as opposed to $1 \mathbin{:} 1$/preemptive scheduling, which would try to make progress on all work, or work-stealing scheduling, which is only depth-first local to a thread and is globally breadth-first).

@@

>>>>

In $N (k \mathbin{:} 1) \mathbin{:} M$ scheduling, we have $N$ units of work mapped onto $M$ CPU threads. Additionally, each of those units of work may be composed of $k$ cooperatively scheduled tasks. This is achieved by pinning the $k$ tasks in a group to one CPU thread, while load balancing units of work across the available cores using the \verb|partr| scheduler, a novel implementation of a parallel depth-first scheduling algorithm. This will discussed future in section \ref{subsub:implementation}.

@@

\begin{lstlisting}[language = Julia]
# perform a merge sort on `v` using parallel threads
function psort(v::AbstractVector)
    hi = length(v)
    if hi < 100_000 # below some cutoff, run in serial
        return sort(view(v, 1:hi), alg = MergeSort)
    end
# split the range and sort the halves in parallel recursively
    mid = (1 + hi) >>> 1
    half = Threads.@spawn psort(view(v, 1:mid))
    right = psort(view(v, (mid + 1):hi))
    left = fetch(half)::typeof(right)
# perform the merge on the result
    out = similar(v)
    merge!(left, right, out)
    return out
end

function merge!(left, right, out)
    ll, lr = length(left), length(right)
    @assert ll + lr == length(out)
    i, il, ir = 1, 1, 1
    @inbounds while il <= ll && ir <= lr
        l, r = left[il], right[ir]
        if isless(r, l)
            out[i] = r
            ir += 1
        else
            out[i] = l
            il += 1
        end
        i += 1
    end
    @inbounds while il <= ll
        out[i] = left[il]
        il += 1
        i += 1
    end
    @inbounds while ir <= lr
        out[i] = right[ir]
        ir += 1
        i += 1
    end
    return out
end
\end{lstlisting}

@@

The algorithm given here is limited in the theoretical scaling capability, since the merge step is not parallelized. On large core counts, that can be important, so please see our supplementary code in \ref{subsub:pmerge}.

@@

\begin{lstlisting}[language = Julia, inputencoding = utf8, literate = {⊕}{{$\oplus$}}1]
using .Threads: @threads
function prefix_threads!(⊕, y::AbstractVector)
    l = length(y)
    k = ceil(Int, log2(l))
    # do reduce phase
    for j = 1:k
        @threads for i = 2^j:2^j:min(l, 2^k)
            @inbounds y[i] = y[i - 2^(j - 1)] ⊕ y[i]
        end
    end
    # do expand phase
    for j = (k - 1):-1:1
        @threads for i = 3*2^(j - 1):2^j:min(l, 2^k)
            @inbounds y[i] = y[i - 2^(j - 1)] ⊕ y[i]
        end
    end
    return y
end

@@

Changing the I/O code (files, streams, folders, and other platform code) to work on from any thread was another big project. The existing design requires an underlying asynchronous library, with a design similar to Windows IOCP, to efficiently manage large numbers of open event sources and provide the simplicity and concision of the logic shown in section \ref{subsub:background} on all platforms. For this, we have been using the libuv library. This lets us have most platform-specific code isolated in a separately tested library and provide more commonality in our runtime library. As an initial implementation to make this library safe to use from threads, we've used one big IO lock around all calls to it. However, this library also has callbacks and will block to wait for external incoming events, so we also needed to integrate it fully with the task scheduler to get it to cooperatively release the lock on demand. We were able to do so by adding an asynchronous channel (\verb|uv_async_t|) to wake the one thread running the event loop while all other threads sleep on a system mutex (\verb|uv_cond_t|) when there is no work for them to perform. When try entering the event loop, we do so only if the count of currently waiting tasks is zero. In the future, this work may allow us to move the event loop entirely to a separate thread (and/or multiple threads). It seems that this design change may thus be making threading support a mandatory requirement for the underlying VM—with the advantage we that we can get more throughput on the large-core systems that are only becoming more common.

@@

A prototype implementation of the \verb|partr| scheduler was first written for us in C by Kiran Pamnany of Intel back in late 2016\footnote{\verb|partr| codebase: \url{https://github.com/kpamnany/partr}}, following research done on scheduling threads for beneficial cache sharing for best throughput\cite{Chen:2007:STC:1248377.1248396}. The goal of this work was effortless composition of threading-capable libraries with a globally depth-first work ordering (as opposed to $1 \mathbin{:} 1$/preemptive scheduling, which would try to make progress on all work, or work-stealing scheduling, which is only depth-first local to a thread and is globally breadth-first).

@@

\appendix
\section{Parallel Merge}
\label{subsub:pmerge}

For doing the merge step of a merge sort in parallel, we only need to add a bit of code to the \verb|merge!| function. Let's call the result \verb|`merge!|. The details of this algorithm can be found elsewhere, although most textbooks do not give the stable algorithm used here. This is a minor enhancement and simplification of the algorithm given in the original paper\cite{Gavril:1975:MPP:361020.361216}, as described in ``Structured Parallel Programming"\cite{McCool:2012:SPP:2385466} in section 13.1 on page 300, and in a later \href{https://software.intel.com/en-us/articles/a-parallel-stable-sort-using-c11-for-tbb-cilk-plus-and-openmp}{blog post by one of the authors}\cite{parallel-merge-intel}.

\begin{lstlisting}[language = Julia, inputencoding = utf8, literate = {÷}{{$\div$}}1]
function pmerge!(left, right, out)
    ll, lr = length(left), length(right)
    @assert ll + lr == length(out)
    if length(out) < 100_000
        # below some threshold, just do the merge
        merge!(left, right, out)
    else
        # split the larger chunk in half, then binary search the
        # smaller half to split it
        if ll > lr
            jl = ll ÷ 2
            # stable sort: find the last entry in right
            # strictly smaller than l
            jr = searchsortedfirst(right, left[jl]) - 1
        else
            jr = lr ÷ 2
            # stable sort: find the last entry in left not bigger
            # than r
            jl = searchsortedlast(left, right[jr])
        end
        @sync begin
            let left = view(left, 1:jl),
                right = view(right, 1:jr),
                out = view(out, 1:(jl + jr))
                Threads.@spawn pmerge!(left, right, out)
            end
            let left = view(left, (jl + 1):ll),
                right = view(right, (jr + 1):lr),
                out = view(out, (jl + jr + 1):length(out))
                pmerge!(left, right, out)
           end
        end
    end
    nothing
end
\end{lstlisting}

This was tested with 8 hyperthreads on the same 4 core laptop as before on a vector \verb|a| consisting of 20M random doubles (Float64 elements). With the parallel \verb|pmerge!|, \verb|@btime psort(a)| reported a runtime of 533 ms vs. 743 ms for the serial merge \verb|merge!|!
